Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
Trim intro

Dynamically cold stellar streams are formed during the tidal disruption of stellar
systems by the gravitational field of their host galaxy.
The phase-space density and mean track of stars in streams therefore encode
information about the underlying distribution and shape of mass on galactic
scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
% Most stellar streams are found in the halos of galaxies and are therefore
% important tracers of the large-scale distribution of dark matter in galactic
% halos.
% Long, thin stellar streams are also excellent laboratories for studying
% small-scale structures in the mass distribution:
% thin streams can remain coherent for tens of orbital periods, depending in
% detail on the symmetries of the underlying galactic mass distribution (e.g.,
% \citealt{Erkal:2016a}).
Being dynamically cold, stellar streams can retain any imprints on its phase-space density from encounters with a massive perturber for close to its total survival time \citep[e.g.,][]{Yoon:2011}.
% Encounters between massive perturbers and stream stars imprint variations in the
% phase-space density that can persist and remain visible for close to their total
% survival time (e.g., \citealt{Yoon:2011}).
In this sense, streams represent one of the most promising directions for
testing the existence of small-scale dark matter sub-halos, predicted by
standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).

Over 30 candidate stellar streams have been discovered in the Milky Way \citep[see][for a recent review]{Grillmair:2016, Newberg:2016}.
% thanks to large-area, multi-band photometric surveys (especially
% the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
% The Milky Way streams have a wide range of properties such as length, velocity
% dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
% Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
% also the recent review \citealt{Grillmair:2016, Newberg:2016}).
% , and the majority
% have no obvious, bound progenitor systems;
% of the thin, likely globular-cluster-origin streams, only two have clear
% progenitors (NGC 5466 and Palomar 5).
% Most streams were first detected as a linear overdensity in stellar counts filtered to preferentially select faint blue stars, most likely members of a former Galactic satellite.
GD-1 is the most prominent among the thin streams, discovered as an arcing overdensity of faint blue stars in the SDSS photometry \citep{Grillmair:2006}.
It spans $\sim 60^\circ$ on the sky, which at its distance between $\sim 7$--$10~\textrm{kpc}$ translates to a physical length of $\sim 15~\kpc$.
No remnant progenitor has been associated with the stream, but its
width ($\sigma \approx 12'$), metallicity ($\feh \approx -1.4$), and estimated
stellar mass ($M \approx 2 \times 10^4~\msun$)
imply that GD-1 is a disrupted globular cluster \citep{Koposov:2010}.

Its length and location in the
Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make GD-1 an
ideal object for constraining the gravitational potential of dark matter in the
inner Milky Way.
Indeed, both from fitting orbits to binned phase-space measurements along the
GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
phase-space coordinates \citep{Bowden:2015, Bovy:2016}, we learned that the
dark matter distribution in the inner $\approx20\,\rm kpc$ is consistent with being close to
spherical.
This is in line with findings from simulated Milky Way-like galaxies; while dark matter halos are in general expected to be triaxial \citep{Allgood:2006}, the baryonic processes forming galaxies tend to make their inner regions more spherical \citep{Zhu:2016}.

An implied old dynamical age, combined with its orbital properties, makes GD-1 a prime stream to not only map the dark matter on large scales, but to also study its small scale structure.
Density variations in the stream can be induced from interactions with dark matter subhalos \citep[e.g.,][]{Ngan:2014}, but also from interactions with giant molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar \citep{Pearson:2017}.
However, these baryonic effects are unlikely to have perturbed GD-1 because of its large pericentric distance and retrograde orbit with respect to
the Galactic bar.
By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
mass of $10^5~\msun$, and by assuming properties of the subhalo population from dark matter-only simulations \citep[scaled to the Milky Way halo mass of $10^{12}\,\msun$, and reduced in number by a factor of 3 to account for disruption by the Galactic disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
($\approx 5$--$7^\circ$) gap caused by an interaction with a
$10^6$--$10^7~\msun$ subhalo.
A definitive ruling on the presence of dark matter subhalos in this mass regime would directly inform on the nature of dark matter \citep[e.g.][]{Bullock:2017}.

Photometric studies of GD-1 revealed the presence of density variations and gaps along the stream, as well as wiggles in the main stream track \citep[][]{Carlberg:2013, DeBoer:2018}.
Purely photometric stream studies suffer to some extent from contamination of the Milky Way foreground, whose small-scale spatial variations could partially account for density inhomogeneities reported along cold streams \citep[e.g.,][]{Ibata:2016}.
In this \article, we improve upon the selection of likely GD-1 members by using astrometric data from the \gaia\ mission, in addition to precise photometry from the \pans\ survey (\S\,\ref{sec:data}).
This data enabled us to measure global properties of GD-1 to an unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures of stream stars beyond the main track (\S\,\ref{sec:res_gap}).
Implications of this revolutionary remapping of GD-1 are discussed in \S\,\ref{sec:discussion}.
% \todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
% (\citealt{Carlberg:2013}).
% More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
% interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2018}):
% [summarize their conclusions].
% \todo{these analyses relied on binned data, still contaminated by MW stars}
%
% In this \article, we present clearest view of the GD-1 stream to date using
% astrometric data from the \gaia\ mission data release 2 (\DR) combined with
% precise photometry from the \pans\ survey.
% We find XX under-densities gaps, some consistent with ...
% We revisit...do not try to model or fit for Galactic potential.
% Focus on stream properties, data.

>>>>

Stellar streams are formed during the tidal disruption of stellar systems by the
gravitational field of their host galaxy.
The phase-space density and mean track of stars in streams therefore encode
information about the underlying distribution and shape of mass on galactic
scales (e.g., \citealt{Johnston:1999, Bonaca:2018}).
% Most stellar streams are found in the halos of galaxies and are therefore
% important tracers of the large-scale distribution of dark matter in galactic
% halos.
% Long, thin stellar streams are also excellent laboratories for studying
% small-scale structures in the mass distribution:
% thin streams can remain coherent for tens of orbital periods, depending in
% detail on the symmetries of the underlying galactic mass distribution (e.g.,
% \citealt{Erkal:2016a}).
Dynamically-cold stellar streams, formed from disrupted globular clusters, can
also retain imprints on their phase-space density from encounters with a massive
perturber for close to their total survival time \citep[e.g.,][]{Yoon:2011}.
% Encounters between massive perturbers and stream stars imprint variations in the
% phase-space density that can persist and remain visible for close to their total
% survival time (e.g., \citealt{Yoon:2011}).
In this sense, streams represent one of the most promising directions for
testing the existence of small-scale dark matter sub-halos, predicted by
standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).

Over 30 candidate stellar streams have been discovered in the Milky Way \citep[see][for a recent review]{Grillmair:2016, Newberg:2016}.
% thanks to large-area, multi-band photometric surveys (especially
% the Sloan Digital Sky Survey, SDSS; \citealt{York:2000}).
% The Milky Way streams have a wide range of properties such as length, velocity
% dispersion, and density (e.g., \citealt{Odenkirchen:2001, Grillmair:2006,
% Grillmair:2006b, Belokurov:2006, Belokurov:2007, Bonaca:2012, Shipp:2018}; see
% also the recent review \citealt{Grillmair:2016, Newberg:2016}).
% , and the majority
% have no obvious, bound progenitor systems;
% of the thin, likely globular-cluster-origin streams, only two have clear
% progenitors (NGC 5466 and Palomar 5).
% Most streams were first detected as a linear overdensity in stellar counts filtered to preferentially select faint blue stars, most likely members of a former Galactic satellite.
The GD-1 stream is the most prominent among the thin streams, discovered as an
overdensity of faint blue stars in the SDSS photometry \citep{Grillmair:2006}.
GD-1 spans $\sim 60^\circ$ on the sky, which, at its distance ($\sim
7$--$10~\textrm{kpc}$) translates to a physical length of $\sim 15~\kpc$.
No remnant progenitor for the stream has been found, but its width ($\sigma
\approx 12'$), metallicity ($\feh \approx -1.4$), and estimated stellar mass ($M
\approx 2 \times 10^4~\msun$) imply that GD-1 is a disrupted globular cluster
\citep{Koposov:2010}.

Its length and location in the Galactic halo (pericentric distance
$r_\textrm{peri} \sim 13~\kpc$) make GD-1 an ideal object for constraining dark
matter properties in the inner Milky Way.
Both from fitting orbits to binned phase-space measurements along the GD-1
stream (\citealt{Koposov:2010}) and from modeling the stream track in
phase-space coordinates \citep{Bowden:2015, Bovy:2016}, it appears that the dark
matter distribution within $\approx20\,\rm kpc$ is close to spherical.
This is consistent with findings from simulated Milky Way-like galaxies:
simulated dark-matter-only halos tend to be triaxial \citep{Allgood:2006}, but
baryonic processes tend to make their inner regions more spherical
\citep{Zhu:2016}.

Its implied old dynamical age, combined with its orbital properties, makes GD-1
a prime stream to also study the small scale structure of dark matter.
While density variations in streams can be induced from interactions with giant
molecular clouds \citep{Amorisco:2016} or resonant encounters with the bar
\citep{Pearson:2017}, these baryonic effects are unlikely to affect GD-1 because
of its large pericentric distance and retrograde orbit with respect to the
Galactic bar.
Therefore, observed density variations in the GD-1 stream could instead be an
indication of past interactions with dark matter subhalos
\citep[e.g.,][]{Ngan:2014}.
By estimating a stream age of $\approx 4~\textrm{Gyr}$, assuming a progenitor
mass of $10^5~\msun$, and by assuming properties of the subhalo population from
dark matter-only simulations \citep[scaled to the Milky Way halo mass and
reduced in number by a factor of 3 to account for disruption by the Galactic
disk]{Springel:2008, Diemand:2008}, \citet{Erkal:2016} predict that GD-1 could
have up to one significant, wide ($\approx 5$--$7^\circ$) gap caused by an
interaction with a $10^6$--$10^7~\msun$ subhalo.
A definitive ruling on the presence of dark matter subhalos in this mass regime
would directly inform on the nature of dark matter \citep[e.g.][]{Bullock:2017}.

Photometric studies of GD-1 revealed the presence of density variations and gaps
along the stream, as well as wiggles in the main stream track
\citep[][]{Carlberg:2013, DeBoer:2018}.
However, purely photometric studies of streams suffer from contamination of the
Milky Way foreground, whose small-scale spatial variations could partially
account for density inhomogeneities reported along cold streams
\citep[e.g.,][]{Ibata:2016}.
In this \article, we improve upon the selection of likely GD-1 members by using
astrometric data from the \gaia\ mission, in addition to precise photometry from
the \pans\ survey (\S\,\ref{sec:data}).
These data enable us to measure density variations along GD-1 to an
unprecedented precision (\S\,\ref{sec:res_global}), and detect clear signatures
of stream stars beyond the main track (\S\,\ref{sec:res_gap}).
Implications of this revolutionary remapping of GD-1 are discussed in
\S\,\ref{sec:discussion}.
% \todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
% (\citealt{Carlberg:2013}).
% More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
% interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2018}):
% [summarize their conclusions].
% \todo{these analyses relied on binned data, still contaminated by MW stars}
%
% In this \article, we present clearest view of the GD-1 stream to date using
% astrometric data from the \gaia\ mission data release 2 (\DR) combined with
% precise photometry from the \pans\ survey.
% We find XX under-densities gaps, some consistent with ...
% We revisit...do not try to model or fit for Galactic potential.
% Focus on stream properties, data.
